On Optimality of Finite Element Integration

### 2015-10-15 TOMS_Paper/assembly.tex
Several fixes, from the abstract up to the second proof

\begin{abstract}
We tackle the problem of automatically generating optimal finite element integration routines given a high level specification of arbitrary multilinear forms. Optimality is defined in terms of floating point operations required to execute a loop nest. The generation of optimal loop nests is driven by a model that exploits mathematical properties of the domain of interest. A theoretical analysis and extensive experimentation prove the effectiveness of our approach, showing systematic performance improvements over a number of alternative code generation systems.
\end{abstract}

@@

\begin{Def}[Outer product loop nest]
An outer product loop nest is a perfect multilinear loop nest of arity $n \leq 2$ in which all of the expressions appearing in the body of the innermost loop are summations, and every loop index occurs in every summand.
\end{Def}

@@

\subsection{Transformation Space}
To synthesize optimal implementations we need:
\begin{itemize}
\item a characterization of the space of legal loop transformations for the class of loop nests considered
\item a cost model to select the optimal point in the transformation space
\end{itemize}
In this section, we construct the transformation space. We leave a few loose ends, which we progressively tie up in Section~\ref{sec:optimal-synthesis}.

@@

\begin{figure}[h]\begin{CenteredBox}
{\subfigcapskip = 7pt \subfigure[With sharing]{\label{code:multi_loopnest_a}\lstinputlisting[basicstyle=\footnotesize\ttfamily]{listings/multilinear_loopnest.code}}}
~~~~~~~~~~~~~~
{\subfigcapskip = 2pt \subfigure[Optimal form]{\label{code:multi_loopnest_b}\lstinputlisting[basicstyle=\footnotesize\ttfamily]{listings/multilinear_loopnest_opt.code}}}
\end{CenteredBox}\caption{A simple outer product loop nest. Note that $L^{sf}_i = 2$ and $L^{sf}_j = 1$.}\end{figure}

Figure~\ref{code:multi_loopnest_a} shows an example of a trivial outer product loop nest of arity $n=2$ with sharing along dimension $j$ induced by $b_j$. Figure~\ref{code:multi_loopnest_b} shows an optimal synthesis for the loop nest in Figure~\ref{code:multi_loopnest_a}. More in general, arbitrarily complex outer product loop nests can systematically be reduced to optimal form. The constructive proof that we provide below focuses on outer product loop nests of arity $n=2$ (for $n < 2$ the proof is trivial; for $n > 2$, as we elaborate later, a more general treatise would be required).

@@

We now have to demonstrate that the chosen factorization strategy, based on sharing factors, leads to optimality. This reduces to prove that the two following facts are true (we do it by contradiction).
\begin{enumerate}
\item Taking $L_j$, the loop with higher sharing factor, as point of departure would lead to a sub-optimal $\Lambda'$ (optimal only in the best case). If we assumed this were false, there would be two possibilities to consider: A) both $L_i$ and $L_j$ present sharing over disjoint sets of outer products; B) at least one outer product induces sharing along both $L_i$ and $L_j$. Case A) is quite trivial since there are no side effects in swapping the order of factorizations, so $\Lambda'$ would actually be optimal (best case scenario). In case B), due to linearity, factorization produces a summation of $L^{sf}_{j}$ outer products, as opposed to $L^{sf}_{i}$ if we had started with $L_i$. However, note that factorization followed by hoisting is always a winning strategy: evaluation, at every iteration, of hoisted symbols requires one operation (a sum) instead of the two operations (a product and a sum) needed by an outer product. Since $L^{sf}_{j} > L^{sf}_{i}$, it is clear that $\Lambda'$ could not be optimal, which contradicts the hypothesis.
\item Partial elimination of sharing would lead to a sub-optimal $\Lambda'$ (optimal only in the best case). Let us assume, for a moment, that this is false. In this scenario, at least one symbol is excluded from the factored product. Without loss of generality, consider, for instance, $a_i(b_j + c_j + ...) + a_i e_j + ...$. It is fairly obvious to see that either there is another factorization opportunity along $L_j$ ($e_j$ is shared with some other terms), in which case no additional operations need be performed, or $\Lambda'$ is sub-optimal. Consequently, the hypothesis is generally contradicted.
\end{enumerate}

Note that if we were considering outer product loop nests of arity $n > 2$ we would need a much more complex construction to guarantee optimality. 
\end{proof}

We now observe that, for the larger class of finite element integration loop nests, the presence of sharing is a \textit{sufficient but not necessary} condition for being in \textit{non-optimal} form. Consider again the bilinear form implementation in Figure~\ref{code:loopnest}. We pose the following question: are we able to identify sub-expressions within $F$ for which the reduction imposed by $L_i$ can be pre-evaluated, thus obtaining a decrease in operation count proportional to the size of $L_i$, $M$? The transformation we look for is exemplified in Figures~\ref{code:loopnest_red} (input) and~\ref{code:loopnest_nored} (output); Figure~\ref{code:loopnest_red} can be seen as a simple instance of the abstract loop nest in Figure~\ref{code:loopnest}.

\begin{figure}[h]\begin{CenteredBox}
{\subfigcapskip = 19pt \subfigure[With reduction]{\label{code:loopnest_red}\lstinputlisting[basicstyle=\footnotesize\ttfamily]{listings/loopnest_red.code}}}
~~~~~~~~~~
{\subfigcapskip = 2pt \subfigure[Pre-evaluated reduction]{\label{code:loopnest_nored}\lstinputlisting[basicstyle=\footnotesize\ttfamily]{listings/loopnest_nored.code}}}
\end{CenteredBox}\caption{Exposition (through factorization) and pre-evaluation of a reduction.}\end{figure}

Pre-evaluation opportunities can be exposed through an exploration of the expression tree transformation space. For arbitrary loop nests, this can be challenging. Further, the following issues arise when considering pre-evaluation opportunities:
\begin{itemize}
\item as opposed to what happens with hoisting in outer product loop nests, the temporary variable size is proportional to the number of non-reduction loops crossed (for the bilinear form implementation in Figure~\ref{code:loopnest}, $N \cdot O$ for sub-expressions depending on $[L_i, L_j, L_k]$ and $L \cdot N \cdot O$ for those depending on $[L_e, L_i, L_j, L_k] $). This might shift the loop nest from a CPU-bound to a memory-bound regime, which might be counter-productive for actual runtime
\item the transformations exposing pre-evaluation opportunities could increase the arithmetic complexity (e.g., expansion may increase the operation count; further examples will be provided later). This could overwhelm the gain inherent in pre-evaluation.
\end{itemize}

To summarize, so far we have highlighted four problems:
\begin{enumerate}
\item The need for an algorithm to expose pre-evaluation opportunities
\item Potential explosion in working set size
\item Potential increase in operation count due to manipulation of the expression tree
\item The need for a strategy to coordinate sharing elimination and pre-evaluation opportunities (sharing elimination could inhibit pre-evaluation; pre-evaluation might expose further sharing)
\end{enumerate}

@@

\subsection{Memory constraints}
\label{sec:mem-const}
We will first consider two examples which motivate the need for memory constraints. Our point of departure is again the bilinear form implementation in Figure~\ref{code:loopnest}. Analogous considerations apply to forms of arbitrary arity.

@@

A second, more general observation is that, for certain forms and discretizations, aggressive hoisting can make the working set exceed the size of some level of local memory (e.g. the last level of private cache on a conventional CPU, the shared memory on a GPU). For example, pre-evaluating geometry-independent expressions outside of $\Lambda$ requires temporary arrays of size $N\cdot O$ for bilinear forms and of size $N$ (or $O$) for linear forms. This can sometimes break such a ``local memory threshold''. In our experiments (Section~\ref{sec:perf-results-forms}) we will carefully study this aspect.

Based on these considerations, we establish the following set of memory constraints:
\begin{enumerate}
\item the size of a temporary due to code motion cannot be larger than that of the outer product loop nest iteration space;
\item the total amount of memory occupied by the temporaries due to code motion cannot exceed a certain threshold, \texttt{$T_H$}.
\end{enumerate}
A corollary of constraint 1) is that hoisting expressions involving geometry-dependent terms outside of $\Lambda$, which is a possibility we discussed at the beginning of this section, is now forbidden. This clearly shrinks the transformation space, pruning points that most likely would result in sub-optimal performance.

@@

To partly answer this question, we make use of a result -- the foundation of tensor contraction mode -- from~\citeN{Kirby:TC}. Essentially, multilinear forms can be seen as sums of monomials, each monomial being an integral over the equation domain of products (of derivatives) of functions from discrete spaces; such monomials can always be reduced to a product of two tensors (see Section~\ref{sec:background}). We interpret this result at the loop nest level: with an input as in Figure~\ref{code:loopnest}, we can always dissect \texttt{F} into distinct sub-expressions (the monomials). Each sub-expression is factorizable so as to split constants from $[L_i, L_j, L_k]$-dependent terms. These $[L_i, L_j, L_k]$-dependent terms can be hoisted outside of $\Lambda$ and stored into temporaries. As part of this process, the reduction induced by $L_i$ is evaluated. Consequently, $L_i$ disappears from $\Lambda$. This transformation is actually what we were referring to when we introduced pre-evaluation in Section~\ref{sec:optimal-impl}.

@@

\begin{figure}
\begin{lstlisting}[numbers=left, basicstyle=\small\ttfamily, frame=single]
dissect the input expression into monomials
for each monomial $m$:
  if $\theta(m)$ > 0 and memory constraints satisfied:
    mark $m$ as candidate for pre-evaluation
for each monomial $m$:
  if $m$ does not share terms with $m'$, an unmarked monomial:
    extract $m$ into a separate loop nest
    apply pre-evaluation to $m$
for each expression:
  remove sharing    
\end{lstlisting}
\caption{High level view of the transformation algorithm}
\label{code:intuition}
\end{figure}

Our choice is to study the impact of pre-evaluation, as number of operations saved or introduced, ``locally''; that is, for each monomial, ``in isolation''. We then introduce the cost function $\theta : M \rightarrow \mathbb{Z}$, where $M$ is the set of monomials in the form. $\theta(m)$ represents the change in operation count if pre-evaluation were applied to monomial $m$: a negative value indicates an increase, whereas a positive value indicates a gain. If pre-evaluation is found profitable (i.e., $\theta(m) > 0$), the sub-expression implementing $m$ is extracted; the surrounding loop nest is suitably ``cloned''; a sequence of transformation steps -- involving expansion, factorization, and code motion -- takes place (details in Section~\ref{sec:codegen}); and, finally, the hoisted code is pre-evaluated using symbolic execution. The result is a set of $n$-dimensional tables (these can be seen as ``slices'' of the reference tensor at the math level), $n$ being the arity of the multilinear form.

Because of symmetries in basis functions, pre-evaluation could produce identical tables. These are mapped to the same temporary. The transformed sub-expression (monomial) can therefore be characterized by the presence of sharing. This is why the last two lines of the algorithm, which goal is to eliminate sharing following the procedure described in Proposition~\ref{prop:optimal-mln}, are transparently applied to all sub-expressions, regardless of whether pre-evaluation was used. The output of the transformation algorithm is as in Figure~\ref{code:loopnest-opt}, assuming the usual input of Figure~\ref{code:loopnest}.

@@

Note that due to the memory constraints imposed in the previous section, sharing elimination and pre-evaluation fully describe the transformation space in which the optimal is searched. We have defined them as high level operations on top of traditional expression-restructuring transformations such as code motion, re-association, and factorization. Other classes of loop transformations, for instance aimed at improving data locality or exposing parallelism, can obviously be neglected since they do not impact the operation count.

@@

\begin{Prop}
\label{prop:optimal-approach}
Consider an arbitrary multilinear form and the corresponding loop nest $\Lambda$ implementing it. The form comprises a set of monomials, $M$. Let $P$ be the set of pre-evaluated monomials, and let $Z$ be the set of non pre-evaluated monomials (i.e., $Z = M \setminus P$). Assume that:
\begin{enumerate}
\item pre-evaluating different monomials does not result in identical tables;
\item there is no sharing of terms between monomials in $P$ and monomials in $Z$.
\end{enumerate} 
Then, the transformation algorithm produces an optimal $\Lambda'$ as established by Definition~\ref{def:optimality}.
\end{Prop}
\begin{proof}
We first comment on the assumptions. First, we notice that neither of them is a sufficient condition for non-optimality. In addition: 1) is inherently a pathological case, which we neglect; 2) might actually occur in forms with several monomials, although, in practice, this was never the case in none of the problems that we have investigated (which are more than those studied in this article). Therefore, we also neglect this scenario, which would otherwise require a ``global analysis'' of $M$.

The operation count for $\Lambda'$ is $\Lambda'_{ops} = L N O (\Lambda'_{ops_1} + I \Lambda'_{ops_2})$ where $\Lambda'_{ops_1}$ and $\Lambda'_{ops_2}$ represent the operation counts for $P$ and $Z$, respectively. $L$, $M$, $N$, and $O$ are the trip counts of the corresponding loops.

@@

A) By pre-evaluating the monomials in $Z_P$, we clearly have $\Lambda''_{ops_2} = \Lambda'_{ops_2} - I \cdot\delta$, where $\delta > 0$ is a parameter to indicate the reduction in operation count exhibited by the non pre-evaluated monomials. To prove A), we then need to convince ourselves that $\Lambda''_{ops_1} -  I \cdot\delta \geq \Lambda'_{ops_1}$ holds. The only (corner) situation in which this might occur is whenever some of the pre-evaluated monomials in $Z_P$ result in tables identical to those already deriving from $P$. If that were the case, more sharing elimination opportunities would be exposed, which in turn could break the optimality claim. Assumption 1) prevents this situation, so the statement is true.

B) Let us consider the non-trivial case: there are at least two monomials $p$ and $z$ that share terms, with $p \in P_Z$ and $z \in Z$. We have to consider the fact that there could be (practically rare) circumstances in which despite $\theta_{wo}(z) < \theta_w(z)$ and $\theta_{w}(p) < \theta_{wo}(p)$, we could have $\theta^G_{wo}(z, p)  < \theta_w(p) + \theta_{wo}(z)$, where $\theta^G_{wo}$ is a hypothetical global oracle. The intuition behind this is that there could be a larger potential gain in sharing elimination that goes lost if monomials are inspected in isolation. An extreme example is a summation of mass matrices, in which each monomial is characterized by same test and trial functions, and by a different pre-multiplying coefficient (i.e., known) function. The previously discussed assumption 2) allows us to neglect this possibility, which then makes B) true. 
\end{proof}

\subsection{Cost Function}
\label{sec:op_count}
It remains to tie up one loose end: the definition of the cost function $\theta$. Since $\theta$ is expected to be used by a compiler to drive the transformation algorithm, requirements are simplicity and velocity.

We define $\theta(m) = \theta_{wo}(m) - \theta_{w}(m)$, where $\theta_{wo}$ and $\theta_{w}$ represent the operation counts for a monomial $m$ without and with pre-evaluation, respectively. In the following, we explain how to derive these two values.

>>>>

\begin{abstract}
We tackle the problem of automatically generating optimal finite element integration routines given a high level specification of arbitrary multilinear forms. Optimality is defined in terms of floating point operations given a memory bound. We provide an approach to explore the transformation space of loops and expressions typical of integration routines and discuss the conditions for optimality. A theoretical analysis and extensive experimentation, which shows systematic performance improvements over a number of alternative code generation systems, validate the approach.
\end{abstract}

@@

\begin{Def}[Outer product loop nest]
An outer product loop nest is a perfect multilinear loop nest of arity $n \leq 2$ in which all of the expressions appearing in the body of the innermost loop are summations, and every loop index occurs once in every summand.
\end{Def}

@@

\subsection{Transformation Space}
To synthesize optimal implementations we need:
\begin{itemize}
\item a characterization of the space of legal transformations for the class of loop nests considered impacting the operation count
\item a cost model to select the optimal point in the transformation space
\end{itemize}
In this section, we construct the transformation space. We leave a few loose ends, which we progressively tie up in Section~\ref{sec:optimal-synthesis}.

@@

\begin{figure}[h]\begin{CenteredBox}
{\subfigcapskip = 7pt \subfigure[With sharing. Note that $L^{sf}_i = 2$ and $L^{sf}_j = 1$.]{\label{code:multi_loopnest_a}\lstinputlisting[basicstyle=\footnotesize\ttfamily]{listings/multilinear_loopnest.code}}}
~~~~~~~~~~~~~~
{\subfigcapskip = 2pt \subfigure[Optimal form]{\label{code:multi_loopnest_b}\lstinputlisting[basicstyle=\footnotesize\ttfamily]{listings/multilinear_loopnest_opt.code}}}
\end{CenteredBox}\caption{Reducing a simple outer product loop nest to optimal form.}\end{figure}

Figure~\ref{code:multi_loopnest_a} shows an example of a trivial outer product loop nest of arity $n=2$ with sharing along dimension $j$ induced by $b_j$. Figure~\ref{code:multi_loopnest_b} shows an optimal synthesis for the loop nest in Figure~\ref{code:multi_loopnest_a}. More in general, arbitrarily complex outer product loop nests can systematically be reduced to optimal form. The constructive proof that we provide below focuses on outer product loop nests of arity $n=2$. Note that for $n < 2$ the proof is trivial, whereas for $n > 2$ a more general treatise would be required; fortunately, this case is never of interest in finite element integration.

@@

We now have to demonstrate that the chosen factorization strategy, based on sharing factors, leads to optimality. This reduces to prove that the two following facts are true (we do it by contradiction).
\begin{enumerate}
\item Taking $L_j$, the loop with higher sharing factor, as point of departure would lead to a sub-optimal $\Lambda'$ (optimal only in the best case). If we assumed this were false, there would be two possibilities to consider: A) both $L_i$ and $L_j$ present sharing over disjoint sets of outer products; B) at least one outer product induces sharing along both $L_i$ and $L_j$. Case A) is quite trivial since there are no side effects in swapping the order of factorizations, so $\Lambda'$ would actually be optimal (best case scenario). In case B), due to linearity, factorization produces a summation of $L^{sf}_{j}$ outer products, as opposed to $L^{sf}_{i}$ if we had started with $L_i$. However, note that factorization followed by hoisting is always a winning strategy: evaluation, at every iteration, of hoisted symbols requires one operation (a sum) instead of the two operations (a product and a sum) needed by an outer product. Since $L^{sf}_{j} > L^{sf}_{i}$, it is clear that $\Lambda'$ could not be optimal, which contradicts the hypothesis.
\item Partial elimination of sharing would lead to a sub-optimal $\Lambda'$ (optimal only in the best case). Let us assume, for a moment, that this is false. In this scenario, at least one symbol is excluded from the factored product. Without loss of generality, consider, for instance, $a_i(b_j + c_j + ...) + a_i e_j + ...$. It is fairly obvious to see that either there is another factorization opportunity along $L_j$ ($e_j$ is shared with some other terms), in which case no additional operations need be performed, or $\Lambda'$ is sub-optimal. Consequently, the hypothesis is generally contradicted.
\end{enumerate} 
\end{proof}

We observe, however, that for finite element integration loop nests the presence of sharing is a \textit{sufficient but not necessary} condition for being in \textit{non-optimal} form. Consider again the bilinear form implementation in Figure~\ref{code:loopnest}. We pose the following question: are we able to identify sub-expressions within $F$ for which the reduction imposed by $L_i$ can be pre-evaluated, thus obtaining a decrease in operation count proportional to the size of $L_i$, $M$? The transformation we look for is exemplified in Figures~\ref{code:loopnest_red} and~\ref{code:loopnest_nored}; Figure~\ref{code:loopnest_red}, which represents the input, can be seen as a simple instance of the abstract loop nest in Figure~\ref{code:loopnest}.

\begin{figure}[h]\begin{CenteredBox}
{\subfigcapskip = 19pt \subfigure[With reduction]{\label{code:loopnest_red}\lstinputlisting[basicstyle=\footnotesize\ttfamily]{listings/loopnest_red.code}}}
~~~~~~~~~~
{\subfigcapskip = 2pt \subfigure[Pre-evaluated reduction]{\label{code:loopnest_nored}\lstinputlisting[basicstyle=\footnotesize\ttfamily]{listings/loopnest_nored.code}}}
\end{CenteredBox}\caption{Exposing (through factorization) and pre-evaluating a reduction.}\end{figure}

Pre-evaluation opportunities can be exposed through an exploration of the expression tree transformation space. This would be challenging if we were to deal with arbitrary loop nests and expressions, while we will see that a simpler algorithm can be adopted for the class arising in finite element integration. In any case, however, the following issues should be considered:
\begin{itemize}
\item as opposed to what happens with hoisting in outer product loop nests, the temporary variable size is proportional to the number of non-reduction loops crossed (for the bilinear form implementation in Figure~\ref{code:loopnest}, $N O$ for sub-expressions depending on $[L_i, L_j, L_k]$ and $L N O$ for those depending on $[L_e, L_i, L_j, L_k] $). This might shift the loop nest from a CPU-bound to a memory-bound regime, which might be counter-productive for actual runtime
\item the transformations exposing pre-evaluation opportunities could increase the arithmetic complexity (e.g., expansion may increase the operation count; further examples will be provided later). This could overwhelm the gain inherent in pre-evaluation.
\end{itemize}

To summarize, so far we have highlighted four problems:
\begin{enumerate}
\item The need for an algorithm to expose pre-evaluation opportunities
\item Potential explosion in working set size
\item Potential increase in operation count due to manipulation of the expression tree
\item The need for a strategy to coordinate sharing elimination and pre-evaluation opportunities: sharing elimination could inhibit pre-evaluation, while pre-evaluation might expose further sharing
\end{enumerate}

@@

\subsection{Memory constraints}
\label{sec:mem-const}
Two observations motivate the need for memory constraints. Our point of departure is again the bilinear form implementation in Figure~\ref{code:loopnest}. Analogous considerations apply to forms of arbitrary arity.

@@

The second observation is that, for certain forms and discretizations, aggressive hoisting can make the working set exceed the size of some level of local memory (e.g. the last level of private cache on a conventional CPU, the shared memory on a GPU). For example, pre-evaluating geometry-independent expressions outside of $\Lambda$ requires temporary arrays of size $N O$ for bilinear forms and of size $N$ (or $O$) for linear forms. This can sometimes break such a ``local memory threshold''. In our experiments (Section~\ref{sec:perf-results-forms}) we will carefully study this aspect.

Based on these considerations, we establish the following memory constraints:
\begin{enumerate}
\item the size of a temporary due to code motion cannot be larger than that of the outer product loop nest iteration space;
\item the total amount of memory occupied by the temporaries due to code motion cannot exceed a certain threshold, \texttt{$T_H$}.
\end{enumerate}
A corollary of constraint 1) is that hoisting expressions involving geometry-dependent terms outside of $\Lambda$, which is a possibility we discussed at the beginning of this section, is now forbidden. This clearly shrinks the transformation space, pruning points that most likely would result in sub-optimal performance.

@@

To partly answer this question, we make use of a result -- the foundation of tensor contraction mode -- from~\citeN{Kirby:TC}. Essentially, multilinear forms can be seen as sums of monomials, each monomial being an integral over the equation domain of products (of derivatives) of functions from discrete spaces; such monomials can always be reduced to a product of two tensors (see Section~\ref{sec:background}). We interpret this result at the loop nest level: with an input as in Figure~\ref{code:loopnest}, we can always dissect \texttt{F} into distinct sub-expressions (the monomials). Each sub-expression is factorizable so as to split constants from $[L_i, L_j, L_k]$-dependent terms. These $[L_i, L_j, L_k]$-dependent terms can be hoisted outside of $\Lambda$ and stored into temporaries. As part of this process, the reduction induced by $L_i$ is evaluated. Consequently, $L_i$ disappears from $\Lambda$. This transformation is actually what we were referring to when we more informally introduced pre-evaluation in Section~\ref{sec:optimal-impl}.

@@

\begin{figure}
\begin{lstlisting}[numbers=left, basicstyle=\small\ttfamily, frame=single]
dissect the input expression into monomials
for each monomial $m$:
  if $\theta(m)$ > 0 and memory constraints satisfied:
    extract into a separate loop nest
    apply pre-evaluation
for each expression:
  remove sharing    
\end{lstlisting}
\caption{High level view of the transformation algorithm}
\label{code:intuition}
\end{figure}

Our choice is to evaluate the impact of pre-evaluation, as number of operations saved or introduced, ``locally''; that is, for each monomial, ``in isolation''. We then introduce the cost function $\theta : M \rightarrow \mathbb{Z}$, where $M$ is the set of monomials in the form. $\theta(m)$ represents the change in operation count if pre-evaluation were applied to monomial $m$: a negative value indicates an increase, whereas a positive value indicates a gain. If pre-evaluation is found profitable (i.e., $\theta(m) > 0$), the sub-expression implementing $m$ is extracted; the surrounding loop nest is suitably ``cloned''; a sequence of transformation steps -- involving expansion, factorization, and code motion -- takes place (details in Section~\ref{sec:codegen}); and, finally, the hoisted code is pre-evaluated using symbolic execution. The result is a set of $n$-dimensional tables (these can be seen as ``slices'' of the reference tensor at the math level), $n$ being the arity of the multilinear form.

Because of symmetries in basis functions, pre-evaluation could produce identical tables, which would be mapped to the same temporary. The transformed sub-expression (monomial) can therefore be characterized by the presence of sharing. This is why the last two lines of the algorithm, which goal is to eliminate sharing following the procedure described in Proposition~\ref{prop:optimal-mln}, are transparently applied to all sub-expressions, regardless of whether pre-evaluation was used. The output of the transformation algorithm is as in Figure~\ref{code:loopnest-opt}, assuming the usual input of Figure~\ref{code:loopnest}.

@@

Note that due to the memory constraints imposed in the previous section, sharing elimination and pre-evaluation fully describe the transformation space in which the optimal is searched. We consider them high level operators defined on top of expression-restructuring transformations such as code motion, re-association, and factorization. Other classes of loop or expression transformations, for instance aimed at improving data locality or exposing parallelism, can obviously be neglected since they do not impact the operation count.

@@

\begin{Prop}
\label{prop:optimal-approach}
Consider an arbitrary multilinear form and the corresponding loop nest $\Lambda$ implementing it. The form comprises a set of monomials, $M$. Let $P$ be the set of pre-evaluated monomials, and let $Z$ be the set of non pre-evaluated monomials (i.e., $Z = M \setminus P$). Assume that:
\begin{enumerate}
\item pre-evaluating different monomials does not result in identical tables;
\item there is no sharing of terms between monomials in $P$ and monomials in $Z$.
\end{enumerate} 
Then, the transformation algorithm produces an optimal $\Lambda'$ as established by Definition~\ref{def:optimality}.
\end{Prop}
\begin{proof}
We first comment on the assumptions. 1) Inherently a pathological case, so we simply neglect it. 2) Might occur in forms with several monomials, although this was never the case in the problems that we have so far investigated (which are obviously more than those presented in Section~\ref{sec:perf-results}). These assumptions allow us to avoid a ``global analysis'' of $M$.

@@

A) The only corner situation in which this might occur is whenever some of the pre-evaluated monomials in $Z_P$ result in tables identical to those already deriving from $P$. If that were the case, more sharing elimination opportunities would be exposed, which in turn could break the optimality claim. Assumption 1) prevents this situation, so the statement is true.

B) Let $\theta_{wo}(m)$ and $\theta_{w}(m)$ represent, respectively, the operation count for a monomial $m$ without and with pre-evaluation and let us consider the non-trivial case in which two monomials $p \in P_Z$ and $z \in Z$ share at least one term. We have to contemplate the fact that despite $\theta_{wo}(z) < \theta_w(z)$ and $\theta_{w}(p) < \theta_{wo}(p)$, we could have $\theta^G_{wo}(z, p)  < \theta_w(p) + \theta_{wo}(z)$, where $\theta^G_{wo}$ is a hypothetical global oracle. That is, there could be a larger potential gain in sharing elimination that goes lost if monomials are inspected in isolation. Potential examples are forms in which multiple monomials differ only for the number and/or the space of coefficient functions. The previously discussed assumption 2) allows us to neglect this possibility. 
\end{proof}

\subsection{Cost Function}
\label{sec:op_count}
It remains to tie up one loose end: the construction of the cost function $\theta$. Since $\theta$ is expected to be used by a compiler to drive the transformation algorithm, requirements are simplicity and velocity.

According to what established in the previous section, we have $\theta(m) = \theta_{wo}(m) - \theta_{w}(m)$; we recall that $\theta_{wo}$ and $\theta_{w}$ represent the operation counts without and with pre-evaluation. In the following, we explain how to derive these two values.
